A reproducible framework for undergraduate data science research

### 2019-12-05 Rmd/introduction.Rmd
updates to methods

The need to analyze unprecedentedly large volumes of information combined with 
the development of faster and more powerful computers has fueled advances in 
data science methods for big data. We present below a framework for mentored 
undergraduate data science research and our findings from its initial 
implementation for two statistics students. We conclude with lessons learned 
and suggestions for those who wish to reproduce and refine our framework.

Social media data, including tweets from Twitter and posts from Facebook, are 
available through website application product interfaces (APIs). Twitter 
shares, via a streaming API, a sample of approximately one percent of all 
tweets during an API query time period [@tweet_stream]. 
Researchers have studied tweets for a 
variety of purposes, including 
inference of relationships and social networks among users [@lin2011joint]; 
determination of 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
study of rhetoric in recruiting political supporters [@pelled2018little,@wells2016trump]. 
Recognizing the potential utility of tweets for data science research, we created a 
collection of tweets over time by repeated querying of the Twitter streaming API.

@nolan2010computing argue for students to work with real data. 
Working with real data allows students to develop skill not only in statistical analysis,
but also in data transfer from online sources, in data storage, and in using data 
from multiple file formats. In the case of Twitter data, tweets are stored in
Javascript Object Notation (JSON) [@tweet_json,@json].

Mentoring in the work place and in higher education can have many 
benefits, including improving students' development as thinkers and scholars, confidence in their 
own abilities, integration into the campus community, and interest in graduate training 
[@baker2010beyond,@higgins2001reconceptualizing].
A key component of our data science mentoring framework is the emphasis on using real data to answer real 
scientific questions. We believe that this process develops problem-solving skills that 
students will need in their future careers in data science. 
We encouraged the student to articulate a scientific research question, translate that question 
into quantitative and statistical terms, determine which data could be used to address the 
question, acquire the data, analyze data, visualize results, and communicate what they learned.

@@

**Need to explain above paragraph with our two student examples. What types of questions did they  articulate? How did they translate to quant terms? How did they determine data availability? etc.**

@@

During the time when we first implemented our framework, we served as early-career instructors in 
the statistics department at the University of Wisconsin-Madison. One of us (Hanlon) had prior 
experience in mentoring students, while the other (Boehm) had none. Our initial 
conceptualization of mentoring drew heavily on ideas we first encountered in professional 
development 
activities, including the Delta Program's mentoring class (https://delta.wisc.edu) and @handelsman2005entering. 
Professor Erik Nordheim heavily 
influenced our approach to and philosophy of teaching statistics. 
We studied with Professor Nordheim early in our teaching 
careers, and his emphasis on active learning continues to influence our teaching practices.

>>>>

The need to analyze unprecedentedly large volumes of information combined with 
the development of faster and more powerful computers has fueled advances in 
data science methods for big data. Similar causes have led to a
need for greater 
numbers of scientists with quantitative skills. In efforts to enhance training
and mentoring for trainees, we created a program that emphasizes many 
transferable skills that contribute to career success in data science.

We elected to work with social media data. This choice was deliberate. In making 
this decision, we recognized that social media data, such as tweets from 
Twitter, can be acquired with little cost and that there is growing research 
interest in social media in many social science disciplines, including political 
science, communication studies, and sociology. We also anticipated 
that our undergraduate trainees might be intrigued by the possibility of 
analyzing social media data, since many young adults use Facebook, Twitter, 
Instagram, and related sites.

Some social media data, including tweets from Twitter, are 
available through website application product interfaces (APIs). Twitter 
shares, via a streaming API, a sample of approximately one percent of all 
tweets during an API query time period [@tweet_stream].
Researchers have studied tweets for a 
variety of purposes, including 
inference of relationships and social networks among users [@lin2011joint]; 
determination of 
authorship of specific tweets when multiple persons share a single account [@drob]; and 
study of rhetoric in recruiting political supporters [@pelled2018little;@wells2016trump]. 
Recognizing the potential utility of tweets for data science research, we created a 
collection of tweets over time by repeated querying of the Twitter streaming API.

@nolan2010computing argue for students to work with real data. 
Working with real data allows students to develop skill not only in statistical analysis,
but also in data transfer from online sources, in data storage, and in using data 
from multiple file formats. In the case of Twitter data, tweets are stored in
Javascript Object Notation (JSON) [@tweet_json;@json].

Mentoring in the work place and in higher education can have many 
benefits, including improving students' development as thinkers and scholars, confidence in their 
own abilities, integration into the campus community, and interest in graduate training 
[@baker2010beyond;@higgins2001reconceptualizing].
A key component of our data science mentoring framework is the emphasis on using real data to answer real 
scientific questions. We believe that this process develops problem-solving skills that 
students will need in their future careers in data science. 
We encouraged the student to articulate a scientific research question, translate that question 
into quantitative and statistical terms, determine which data could be used to address the 
question, acquire the data, analyze data, visualize results, and communicate what they learned.

@@

@@

During the time when we first implemented our framework, we served as early-career instructors in 
the statistics department at the University of Wisconsin-Madison. One of us (Hanlon) had prior 
experience in mentoring students, while the other (Boehm) had none. Our initial 
conceptualization of mentoring drew heavily on ideas we first encountered in professional 
development 
activities, including the Delta Program's mentoring class (https://delta.wisc.edu) and @handelsman2005entering. 
Professor Erik Nordheim
influenced our approach to and philosophy of teaching statistics. 
We studied with Professor Nordheim early in our teaching 
careers, and his emphasis on backward design and active learning continues to influence our teaching practices.

### 2019-12-05 Rmd/methods.Rmd
updates to methods

@@

@@

After identifying research questions and publicly available data, 
the next step is to decide on informative data visualizations and quantitative analyses. 
Because 
both projects primarily involved exploratory analyses of times series, 
we encouraged students to think about
visualizations that might reveal relationships over time.

### Examples

@@

>>>>

### Research question formulation

@@

### Data acquisition

@@

### Data analysis and visualization

After identifying research questions and publicly available data, 
the next step is to decide on informative data visualizations and quantitative analyses. 
Because 
both projects involved exploratory analyses of times series, 
we encouraged students to think about
visualizations that might reveal relationships over time.

In the case of the event detection project, we e

### Presentation and communication of results

Students presented their research in a variety of settings. 
Each student presented at the annual undergraduate statistics poster session. 
We also encouraged them to present at the annual university-wide 
undergraduate research symposium.

In planning with students for poster and slide presentations, 
we (Hanlon and Boehm) emphasized
the importance of succinctly stating the research question and its scientific 
context. After clarifying the importance of the question, the student could 
proceed with explaining many of the elements that we've described above. Namely, the 
student would discuss the analyzed data and its acquisition while noting any shortcomings or 
biases of the data. For oral presentations, we suggested that 
students cautiously limit discussion of statistical methods, with the caveat that 
they prepare to answer detailed methodological inquiries during the question and 
answer session. 
Our students created powerful data 
visualizations for their projects. Their presentations also included their major results and future research directions.

In efforts to develop student written communication, we encouraged both students to 
prepare a written senior thesis document that detailed their research. In the senior 
thesis, we suggested that the students describe in rigorous detail their statistical 
methods. The rationale for this distinction, relative to the oral presentation, is 
that a reader doesn't have access to a question and answer session, while a poster 
session attendee may freely ask questions of the author.

## Examples

@@

## Develop and practice skills in reproducible research to promote open science
